---
title: "DESeq2_objects_geneAnnotations"
author: "allyson_demerlis"
date: "2023-03-21"
output: html_document
---

This Rmarkdown involves cleaning the counts matrix, creating the metadata files, and creating the DDS objects for the full time-series model. Significant DGE lists were then annotated using the Pdam genome .gff file. 

Note: DDS objects are saved here as .rds files, and then these are imported into all subsequent Rmarkdown files.

# Install and load necessary packages
```{r, install/load packages}
#if you need to install all the packages, you need the package "BiocManager" first.
# if (!require("BiocManager",quietly=TRUE))
#   install.packages("BiocManager")
#Now you can use BiocManager to install pacakges like DESeq2
#BiocManager::install("DEGreport")

library(DESeq2)

library(DEGreport) # needs BiocManager
  #DEGreport requires lasso2, which is not available for newer versions of R. To install lasso2, run library(remotes) (might need to install that package first), then run:
#install_version("lasso2", "1.2-22")

library(tidyverse) 
```


# Importing gene counts matrix and formatting it
```{r}
#reading in txt file as a table with the headers as the header in the original file
countsmatrix <-read.table("../0_input_files/WH_Pdam_original.Rmatrix.txt",header=TRUE)

#removing the Chromosome column from the dataset, and changing the rows to be the gene names
countsmatrix <-countsmatrix %>% 
  dplyr::select(-c("Chr", "Start", "End", "Strand", "Length")) %>% 
  column_to_rownames("Geneid")

#removing hour 5 from the dataset because it only has 1 wounded coral, so it is not comparable 
countsmatrix <- countsmatrix %>% 
  dplyr::select(!X5331C1:X5335Y)

#function found from stack overflow
#tidying column names so they don't include an X anymore (R puts an X in front of columns that start with numbers)
destroyX = function(es) {
  f = es
  for (col in c(1:ncol(f))){ #for each column in dataframe
    if (startsWith(colnames(f)[col], "X") == TRUE)  { #if starts with 'X' ..
      colnames(f)[col] <- substr(colnames(f)[col], 2, 100) #get rid of it
    }
  }
  assign(deparse(substitute(es)), f, inherits = TRUE) #assign corrected data to original name
}

destroyX(countsmatrix)
```


# Import gene feature annotation information
```{r}
genefeatures <- read.delim(file = "../0_input_files/pdam_genome_IDInfo.gff", header = F)
head(genefeatures)
colnames(genefeatures) <- c("IDGeneInfo")
rownames(genefeatures) <- rownames(countsmatrix)

#removing gene name from the second column
gene_name_split <- str_split_fixed(genefeatures$IDGeneInfo, " ", 2)
print(head(gene_name_split))
colnames(gene_name_split)<- c("Gene_ID","Gene_Function")
head(gene_name_split)
gene_name_split_df<-data.frame(gene_name_split)
head(gene_name_split_df)

#convert the Gene_ID column to the row headers
gene_name_split_df %>%
     remove_rownames() %>%
 column_to_rownames(var = 'Gene_ID') %>% 
mutate(Gene_Function = URLdecode(Gene_Function)) -> Pdam_Gene_Names_Info 
```


# DDS Object
```{r}
metadata = data.frame(sample=colnames(countsmatrix),
                condition = stringr::str_detect(pattern = ".*C.*",string = colnames(countsmatrix)),
                hour = stringr::str_replace(pattern = "(.).*",replacement="\\1",string = colnames(countsmatrix)),
 id = stringr::str_replace(pattern=".(...).*",replacement="\\1",string=colnames(countsmatrix)))

#changing TRUE and FALSE for condition to Control and Wounded
metadata$condition[str_detect(metadata$condition,"TRUE")] <- "Control"
metadata$condition[str_detect(metadata$condition,"FALSE")] <- "Wounded"

metadata <- metadata %>% column_to_rownames("sample")

metadata$condition <- as.factor(metadata$condition)
metadata$hour <- as.factor(metadata$hour)

# creating grouping variable

metadata$group <- factor(paste0(metadata$condition, "_", metadata$hour))
metadata$group <- relevel(metadata$group, ref = "Control_0")

#saveRDS(metadata, file = "metadata.rds")
```

# Testing group (condition_hour)
```{r}
dds_group <- DESeqDataSetFromMatrix(countData = countsmatrix, colData = metadata, design = ~group)
# Control_0 is reference level  

#prefiltering recommended by DESeq2 Vignette (removes anything with reads below 10)
keep_group <- rowSums(counts(dds_group)) >= 10
length(keep_group) #26077
dds_group <- dds_group[keep_group,]
length(dds_group) #21267

dds_group <- DESeq(dds_group)
# Save  object to a file
#saveRDS(dds_group, file = "dds_group.rds")

results_0h <- results(dds_group, contrast = c("group", "Wounded_0", "Control_0"), alpha = 0.05)
summary(results_0h, alpha = 0.05) # 2 downregulated
results_1h <- results(dds_group, contrast = c("group", "Wounded_1", "Control_1"), alpha = 0.05)
summary(results_1h, alpha = 0.05) # 4 down
results_2h <- results(dds_group, contrast = c("group", "Wounded_2", "Control_2"), alpha = 0.05)
summary(results_2h, alpha = 0.05) # 1 up, 2 down
results_4h <- results(dds_group, contrast = c("group", "Wounded_4", "Control_4"), alpha = 0.05)
summary(results_4h, alpha = 0.05) # 43 up, 65 down

#annotate results to include Gene Function
annot_0h<-merge(as.data.frame(results_0h),Pdam_Gene_Names_Info,by='row.names',all=TRUE)
annot_0h<-na.omit(annot_0h)
annot_0h %>% 
  filter(padj < 0.05) %>% 
  mutate(group = "Wounded_0 vs Control_0") -> annot_0h_sig

annot_1h<-merge(as.data.frame(results_1h),Pdam_Gene_Names_Info,by='row.names',all=TRUE)
annot_1h<-na.omit(annot_1h)
annot_1h %>% 
  filter(padj < 0.05) %>% 
  mutate(group = "Wounded_1 vs Control_1") -> annot_1h_sig

annot_2h<-merge(as.data.frame(results_2h),Pdam_Gene_Names_Info,by='row.names',all=TRUE)
annot_2h<-na.omit(annot_2h)
annot_2h %>% 
  filter(padj < 0.05) %>% 
  mutate(group = "Wounded_2 vs Control_2") -> annot_2h_sig

annot_4h<-merge(as.data.frame(results_4h),Pdam_Gene_Names_Info,by='row.names',all=TRUE)
annot_4h<-na.omit(annot_4h)
annot_4h %>% 
  filter(padj < 0.05) %>% 
  mutate(group = "Wounded_4 vs Control_4") -> annot_4h_sig

#save annotated significant DEG lists as .csv files
#rbind(annot_0h_sig, annot_1h_sig, annot_2h_sig, annot_4h_sig) %>%
  #write.csv(file = "Annotated_significant_DEG_lists_timepoint_group_contrasts.csv", row.names = FALSE)
```

# Testing factorial design (condition + hour + condition:hour)

```{r}
dds_factorial <- DESeqDataSetFromMatrix(countData = countsmatrix, colData = metadata, design = ~ hour + condition + hour:condition)

#prefiltering recommended by DESeq2 Vignette (removes anything with reads below 10)
keep_factorial <- rowSums(counts(dds_factorial)) >= 10
length(keep_factorial) #26077
dds_factorial <- dds_factorial[keep_factorial,]
length(dds_factorial) #21267

# 1. Which genes respond to injury (overall)?
dds_lrt_injury <- DESeq(dds_factorial, test = "LRT", reduced = ~ hour)
res_injury <- results(dds_lrt_injury, alpha = 0.05)
summary(res_injury) #34 up, 58 down
# Significant = injury-responsive genes
#saveRDS(dds_lrt_injury, file = "dds_lrt_injury.rds")

# 2. Which genes show time-specific injury response?
dds_lrt_interaction <- DESeq(dds_factorial, test = "LRT", 
                             reduced = ~ hour + condition)
res_interaction <- results(dds_lrt_interaction, alpha = 0.05)
summary(res_interaction) # 1 up, 2 down
# Significant = early vs late responders
#saveRDS(dds_lrt_interaction, file = "dds_lrt_interaction.rds")

# 3. Which genes change over time (regardless of injury)?
dds_lrt_time <- DESeq(dds_factorial, test = "LRT", reduced = ~ condition)
res_time <- results(dds_lrt_time, alpha = 0.05)
summary(res_time) # 3677 up, 3963 down
# Significant = temporal dynamics
#saveRDS(dds_lrt_time, file = "dds_lrt_time.rds")
```

annotate 
```{r}
#annotate results to include Gene Function
annot_res_injury<-merge(as.data.frame(res_injury),Pdam_Gene_Names_Info,by='row.names',all=TRUE)
annot_res_injury<-na.omit(annot_res_injury)
annot_res_injury %>% 
  filter(padj < 0.05) %>% 
  mutate(contrast = "Injury") -> annot_res_injury_sig

annot_res_interaction<-merge(as.data.frame(res_interaction),Pdam_Gene_Names_Info,by='row.names',all=TRUE)
annot_res_interaction<-na.omit(annot_res_interaction)
annot_res_interaction %>% 
  filter(padj < 0.05) %>% 
  mutate(contrast = "Interaction") -> annot_res_interaction_sig

annot_res_time<-merge(as.data.frame(res_time),Pdam_Gene_Names_Info,by='row.names',all=TRUE)
annot_res_time<-na.omit(annot_res_time)
annot_res_time %>% 
  filter(padj < 0.05) %>% 
  mutate(contrast = "Time") -> annot_res_time_sig

#save annotated significant DEG lists as .csv files
# rbind(annot_res_injury_sig, annot_res_interaction_sig, annot_res_time_sig) %>%
#   write.csv(file = "Annotated_significant_DEG_lists_lrt_contrasts.csv", row.names = FALSE)
```

# Generate injury-specific genes 

```{r}
sig_w_1v0 <- rownames(
  results(dds_group, contrast = c("group", "Wounded_1", "Wounded_0"))[!is.na(results(dds_group, contrast = c("group", "Wounded_1", "Wounded_0"))$padj) & results(dds_group, contrast = c("group", "Wounded_1", "Wounded_0"))$padj < 0.05, ]
)
length(sig_w_1v0) #1191

sig_w_2v0 <- rownames(
  results(dds_group, contrast = c("group", "Wounded_2", "Wounded_0"))[!is.na(results(dds_group, contrast = c("group", "Wounded_2", "Wounded_0"))$padj) & results(dds_group, contrast = c("group", "Wounded_2", "Wounded_0"))$padj < 0.05, ]
)
length(sig_w_2v0) #3028

sig_w_4v0 <- rownames(
  results(dds_group, contrast = c("group", "Wounded_4", "Wounded_0"))[!is.na(results(dds_group, contrast = c("group", "Wounded_4", "Wounded_0"))$padj) & results(dds_group, contrast = c("group", "Wounded_4", "Wounded_0"))$padj < 0.05, ]
)
length(sig_w_4v0) #2863

# Get control time effects
res_c_1v0 <- results(dds_group, contrast = c("group", "Control_1", "Control_0"))
res_c_2v0 <- results(dds_group, contrast = c("group", "Control_2", "Control_0"))
res_c_4v0 <- results(dds_group, contrast = c("group", "Control_4", "Control_0"))

# Control temporal changes
sig_c_1v0 <- rownames(res_c_1v0[!is.na(res_c_1v0$padj) & res_c_1v0$padj < 0.05, ]) 
length(sig_c_1v0) #1331
sig_c_2v0 <- rownames(res_c_2v0[!is.na(res_c_2v0$padj) & res_c_2v0$padj < 0.05, ])
length(sig_c_2v0) #3366
sig_c_4v0 <- rownames(res_c_4v0[!is.na(res_c_4v0$padj) & res_c_4v0$padj < 0.05, ])
length(sig_c_4v0) #2710

control_time_genes <- unique(c(
  rownames(res_c_1v0[!is.na(res_c_1v0$padj) & res_c_1v0$padj < 0.05, ]),
  rownames(res_c_2v0[!is.na(res_c_2v0$padj) & res_c_2v0$padj < 0.05, ]),
  rownames(res_c_4v0[!is.na(res_c_4v0$padj) & res_c_4v0$padj < 0.05, ])
))

# Get wounded time effects and filter
res_w_1v0 <- results(dds_group, contrast = c("group", "Wounded_1", "Wounded_0"))
res_w_2v0 <- results(dds_group, contrast = c("group", "Wounded_2", "Wounded_0"))
res_w_4v0 <- results(dds_group, contrast = c("group", "Wounded_4", "Wounded_0"))

sig_w_1v0_injury <- setdiff(
  rownames(res_w_1v0[!is.na(res_w_1v0$padj) & res_w_1v0$padj < 0.05, ]),
  control_time_genes
)
length(sig_w_1v0_injury) #295

sig_w_2v0_injury <- setdiff(
  rownames(res_w_2v0[!is.na(res_w_2v0$padj) & res_w_2v0$padj < 0.05, ]),
  control_time_genes
)

sig_w_4v0_injury <- setdiff(
  rownames(res_w_4v0[!is.na(res_w_4v0$padj) & res_w_4v0$padj < 0.05, ]),
  control_time_genes
)

# Summary
cat("Control time-responsive genes:", length(control_time_genes), "\n\n")
cat("Wounded temporal genes (before/after filtering):\n")
cat("1h:", length(sig_w_1v0), "→", length(sig_w_1v0_injury), "\n")
cat("2h:", length(sig_w_2v0), "→", length(sig_w_2v0_injury), "\n")
cat("4h:", length(sig_w_4v0), "→", length(sig_w_4v0_injury), "\n")
# 1h: 1191 → 295 
# 2h: 3028 → 755 
# 4h: 2863 → 937 

annot_sig_w_1v0_injury<-merge(as.data.frame(res_w_1v0[sig_w_1v0_injury,]),Pdam_Gene_Names_Info,by='row.names',all=TRUE)
annot_sig_w_1v0_injury<-na.omit(annot_sig_w_1v0_injury)
annot_sig_w_1v0_injury %>% 
  mutate(contrast = "Wounded 1h vs 0h (injury-filtered)") -> annot_sig_w_1v0_injury_sig

annot_sig_w_1v0_injury_sig %>% 
  filter(log2FoldChange > 0) %>%
  summarise(count = n()) #182

annot_sig_w_1v0_injury_sig %>% 
  filter(log2FoldChange <0) %>%
  summarise(count = n()) #113

annot_sig_w_2v0_injury<-merge(as.data.frame(res_w_2v0[sig_w_2v0_injury,]),Pdam_Gene_Names_Info,by='row.names',all=TRUE)
annot_sig_w_2v0_injury<-na.omit(annot_sig_w_2v0_injury)
annot_sig_w_2v0_injury %>% 
  mutate(contrast = "Wounded 2h vs 0h (injury-filtered)") -> annot_sig_w_2v0_injury_sig

annot_sig_w_2v0_injury_sig %>% 
  filter(log2FoldChange > 0) %>%
  summarise(count = n()) #346

annot_sig_w_2v0_injury_sig %>% 
  filter(log2FoldChange <0) %>%
  summarise(count = n()) #409

annot_sig_w_4v0_injury<-merge(as.data.frame(res_w_4v0[sig_w_4v0_injury,]),Pdam_Gene_Names_Info,by='row.names',all=TRUE)
annot_sig_w_4v0_injury<-na.omit(annot_sig_w_4v0_injury)
annot_sig_w_4v0_injury %>% 
  mutate(contrast = "Wounded 4h vs 0h (injury-filtered)") -> annot_sig_w_4v0_injury_sig

annot_sig_w_4v0_injury_sig %>%
  filter(log2FoldChange > 0) %>%
  summarise(count = n()) #414

annot_sig_w_4v0_injury_sig %>%
  filter(log2FoldChange <0) %>%
  summarise(count = n()) #523

#save annotated significant DEG lists as .csv files
# rbind(annot_sig_w_1v0_injury_sig, annot_sig_w_2v0_injury_sig, annot_sig_w_4v0_injury_sig) %>%
#   write.csv(file = "Annotated_significant_DEG_lists_wounded_injury_filtered_contrasts.csv", row.names = FALSE)
```

