---
title: "GO_analysis"
author: "allyson_demerlis"
date: "2022-10-25"
output: html_document
---

This markdown is for the Gene Ontology (GO) enrichment analysis using the package "topGO" (Alexa et al. 2006). 

Following https://rpubs.com/LeonYu/995707 and [Dr Chrstine Roper's GitHub](https://github.com/ChristineRoper/Mangrove-coral-thermotolerance-retained/blob/main/RNA%20Seq/4-Functional_annotation/Functional_annotation.R).

# load necessary packages
```{r, install/load packages}
library(tidyverse) 
library(DESeq2)
library(topGO) # needs BiocManager
library(GO.db) # needs BiocManager
library(Rgraphviz) # needs BiocManager
library(scales) #for GO dot plot
library(viridis) #for GO dot plot
library(GSEABase)
library(rrvgo)
library(ggalluvial)
```

# Import DDS object
```{r}
dds_group<-readRDS("rds_files/dds_group.rds")
```

# Gene-to-GO objects
```{r}
#Input required *P. damicornis* GO annotation data and construct Gene-to-GO object for custom annotation mapping
GO_geneID<-readMappings(file="../0_input_files/pdam_genome_GOgenes.txt") #Mike made this file

geneID2GO <- inverseList(GO_geneID)
str(head(geneID2GO))

#Generate gene universe and GO universe from Gene-to-GO and GO-to-Gene objects
geneNames <- names(geneID2GO)
str(head(geneNames))

GONames <- names(GO_geneID)
str(head(GONames))
```

# all DEGs genes combined
```{r}
## Background universe
allgenes <- rownames(dds_group)  # or equivalent

## DEG list
injury_genes <- read_csv("../3_Results/Annotated_significant_DEG_lists_wounded_injury_filtered_contrasts.csv")
genes_by_hour <- read_csv("../3_Results/Annotated_significant_DEG_lists_timepoint_group_contrasts.csv")

injury_genes %>% 
  dplyr::rename(gene = Row.names) -> injury_genes

genes_by_hour %>% 
  dplyr::rename(gene = Row.names) -> genes_by_hour

DEG_genes <- unique(c(injury_genes$gene,
                        genes_by_hour$gene))
```


```{r}
## Build topGO geneList
geneList <- as.integer(allgenes %in% DEG_genes)
names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results$weightFisher <- as.numeric(GO_results$weightFisher)
GO_results <- GO_results[GO_results$weightFisher < 0.05, ]

str(GO_results)
```

GOslim
```{r}
#change Class types to ontology so it matches GOslim
GO_results %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs <- GO_results$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results <- bind_rows(all_slim_results)

# View results
head(slim_results)

```


Plots
```{r}
# Prepare data: select top terms per ontology
top_terms <- GO_results %>%
  group_by(Class) %>%
  arrange(weightFisher) %>%
  slice_head(n = 10) %>%  # Top 10 per ontology
  ungroup() %>%
  mutate(
    GeneRatio = Significant / Annotated,  # Calculate gene ratio
    NegLog10P = -log10(weightFisher),
    Term = fct_reorder(Term, NegLog10P)  # Order by significance
  )

# Create dot plot
top_terms %>% 
    mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) %>% 
ggplot(., aes(x = GeneRatio, y = fct_reorder(Term, GeneRatio))) +
  geom_point(aes(size = Significant, color = NegLog10P)) +
  scale_color_gradient(low = "blue", high = "red", 
                       name = "-log10(P-value)") +
  scale_size_continuous(name = "Gene Count", range = c(1, 8)) +
  facet_grid(Ontology ~ ., scales = "free_y", space = "free_y") +
  labs(x = "Gene Ratio",
       y = "") +
  theme_bw() +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 14),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  ) 
```



# separate time points
```{r}
## Background universe
allgenes <- rownames(dds_group)  # or equivalent

## DEG list
injury_genes <- read_csv("../3_Results/Annotated_significant_DEG_lists_wounded_injury_filtered_contrasts.csv")
genes_by_hour <- read_csv("../3_Results/Annotated_significant_DEG_lists_timepoint_group_contrasts.csv")

injury_genes %>% 
  dplyr::rename(gene = Row.names) %>% 
  filter(contrast == "Wounded 1h vs 0h (injury-filtered)") %>% 
  dplyr::select(gene) -> injury_genes_w_1_0 #295 genes

injury_genes %>% 
  dplyr::rename(gene = Row.names) %>%  
  filter(contrast == "Wounded 2h vs 0h (injury-filtered)") %>% 
  dplyr::select(gene) -> injury_genes_w_2_0 #755

injury_genes %>%
  dplyr::rename(gene = Row.names) %>% 
  filter(contrast == "Wounded 4h vs 0h (injury-filtered)") %>% 
  dplyr::select(gene) -> injury_genes_w_4_0 #937

genes_by_hour %>% 
  dplyr::rename(gene = Row.names) %>% 
  filter(group == "Wounded_0 vs Control_0") %>% 
  dplyr::select(gene) -> genes_by_hour_w0_c0 #2

genes_by_hour %>% 
  dplyr::rename(gene = Row.names) %>% 
  filter(group == "Wounded_1 vs Control_1") %>% 
  dplyr::select(gene) -> genes_by_hour_w1_c1 #4

genes_by_hour %>%
  dplyr::rename(gene = Row.names) %>% 
  filter(group == "Wounded_2 vs Control_2") %>% 
  dplyr::select(gene) -> genes_by_hour_w2_c2 #3

genes_by_hour %>%
  dplyr::rename(gene = Row.names) %>% 
  filter(group == "Wounded_4 vs Control_4") %>% 
  dplyr::select(gene) -> genes_by_hour_w4_c4 #108

```


## injury_genes_w_1_0
```{r}
injury_genes %>% 
  dplyr::rename(gene = Row.names) %>% 
  filter(contrast == "Wounded 1h vs 0h (injury-filtered)") %>% 
  filter(log2FoldChange > 0) -> injury_genes_w_1_0_up

injury_genes %>%
  dplyr::rename(gene = Row.names) %>% 
  filter(contrast == "Wounded 1h vs 0h (injury-filtered)") %>% 
  filter(log2FoldChange < 0) -> injury_genes_w_1_0_down

# Upregulated
injury_genes_w_1_0_up_list <- unique(injury_genes_w_1_0_up$gene)

## Build topGO geneList
geneList <- as.integer(allgenes %in% injury_genes_w_1_0_up_list)

names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_injury_genes_w_1_0_up <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_injury_genes_w_1_0_up$weightFisher <- as.numeric(GO_results_injury_genes_w_1_0_up$weightFisher)
GO_results_injury_genes_w_1_0_up <- GO_results_injury_genes_w_1_0_up[GO_results_injury_genes_w_1_0_up$weightFisher < 0.05, ]

str(GO_results_injury_genes_w_1_0_up) # 15 terms

# Downregulated
injury_genes_w_1_0_down_list <- unique(injury_genes_w_1_0_down$gene)

## Build topGO geneList
geneList <- as.integer(allgenes %in% injury_genes_w_1_0_down_list)

names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_injury_genes_w_1_0_down <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_injury_genes_w_1_0_down$weightFisher <- as.numeric(GO_results_injury_genes_w_1_0_down$weightFisher)
GO_results_injury_genes_w_1_0_down <- GO_results_injury_genes_w_1_0_down[GO_results_injury_genes_w_1_0_down$weightFisher < 0.05, ]

str(GO_results_injury_genes_w_1_0_down) # 19 terms


```

### Goslim
```{r}
# Upregulated

#change Class types to ontology so it matches GOslim
GO_results_injury_genes_w_1_0_up %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_injury_genes_w_1_0_up

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_injury_genes_w_1_0_up <- GO_results_injury_genes_w_1_0_up$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_injury_genes_w_1_0_up %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_injury_genes_w_1_0_up <- bind_rows(all_slim_results)

slim_results_GO_results_injury_genes_w_1_0_up %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> slim_results_GO_results_injury_genes_w_1_0_up_edited

# Merge GO_results with GO_slim to add GOTerms column
slim_results_GO_results_injury_genes_w_1_0_up_edited %>%
  full_join(GO_results_injury_genes_w_1_0_up, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) -> GO_results_injury_genes_w_1_0_up_all

# Downregulated 

#change Class types to ontology so it matches GOslim
GO_results_injury_genes_w_1_0_down %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_injury_genes_w_1_0_down

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_injury_genes_w_1_0_down <- GO_results_injury_genes_w_1_0_down$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_injury_genes_w_1_0_down %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_injury_genes_w_1_0_down <- bind_rows(all_slim_results$BP, all_slim_results$MF)

slim_results_GO_results_injury_genes_w_1_0_down %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> slim_results_GO_results_injury_genes_w_1_0_down_edited

# Merge GO_results with GO_slim to add GOTerms column
slim_results_GO_results_injury_genes_w_1_0_down_edited %>%
  full_join(GO_results_injury_genes_w_1_0_down, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) -> GO_results_injury_genes_w_1_0_down_all

GO_results_injury_genes_w_1_0_up_all %>% 
  mutate(Contrast = "Wounded 1h vs Wounded 0h", Direction = "Upregulated") -> GO_results_injury_genes_w_1_0_up_all

GO_results_injury_genes_w_1_0_down_all %>% 
  mutate(Contrast = "Wounded 1h vs Wounded 0h", Direction = "Downregulated") -> GO_results_injury_genes_w_1_0_down_all

# Combine up and down regulated results
GO_results_genes_by_hour_w_1_0_all <- bind_rows(GO_results_injury_genes_w_1_0_up_all, GO_results_injury_genes_w_1_0_down_all)

#write_csv(GO_results_genes_by_hour_w_1_0_all, "GO_results_genes_by_hour_w_1_0_all.csv")
```


## injury_genes_w_2_0
```{r}
injury_genes %>% 
  dplyr::rename(gene = Row.names) %>% 
  filter(contrast == "Wounded 2h vs 0h (injury-filtered)") %>% 
  filter(log2FoldChange > 0) -> injury_genes_w_2_0_up

injury_genes %>%
  dplyr::rename(gene = Row.names) %>% 
  filter(contrast == "Wounded 2h vs 0h (injury-filtered)") %>% 
  filter(log2FoldChange < 0) -> injury_genes_w_2_0_down

# Upregulated

injury_genes_w_2_0_up_list <- unique(injury_genes_w_2_0_up$gene)

## Build topGO geneList
geneList <- as.integer(allgenes %in% injury_genes_w_2_0_up_list)

names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_injury_genes_w_2_0_up <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_injury_genes_w_2_0_up$weightFisher <- as.numeric(GO_results_injury_genes_w_2_0_up$weightFisher)
GO_results_injury_genes_w_2_0_up <- GO_results_injury_genes_w_2_0_up[GO_results_injury_genes_w_2_0_up$weightFisher < 0.05, ]

str(GO_results_injury_genes_w_2_0_up) # 13 terms

# Downregulated

injury_genes_w_2_0_down_list <- unique(injury_genes_w_2_0_down$gene)

## Build topGO geneList
geneList <- as.integer(allgenes %in% injury_genes_w_2_0_down_list)

names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_injury_genes_w_2_0_down <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_injury_genes_w_2_0_down$weightFisher <- as.numeric(GO_results_injury_genes_w_2_0_down$weightFisher)
GO_results_injury_genes_w_2_0_down <- GO_results_injury_genes_w_2_0_down[GO_results_injury_genes_w_2_0_down$weightFisher < 0.05, ]

str(GO_results_injury_genes_w_2_0_down) # 23 terms
```


### Goslim
```{r}
# Upregulated

#change Class types to ontology so it matches GOslim
GO_results_injury_genes_w_2_0_up %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_injury_genes_w_2_0_up

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_injury_genes_w_2_0_up <- GO_results_injury_genes_w_2_0_up$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_injury_genes_w_2_0_up %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_injury_genes_w_2_0_up <- bind_rows(all_slim_results)

slim_results_GO_results_injury_genes_w_2_0_up %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> slim_results_GO_results_injury_genes_w_2_0_up_edited

# Merge GO_results with GO_slim to add GOTerms column
slim_results_GO_results_injury_genes_w_2_0_up_edited %>%
  full_join(GO_results_injury_genes_w_2_0_up, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) -> GO_results_injury_genes_w_2_0_up_all

# Downregulated 

#change Class types to ontology so it matches GOslim
GO_results_injury_genes_w_2_0_down %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_injury_genes_w_2_0_down

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_injury_genes_w_2_0_down <- GO_results_injury_genes_w_2_0_down$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_injury_genes_w_2_0_down %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_injury_genes_w_2_0_down <- bind_rows(all_slim_results)

slim_results_GO_results_injury_genes_w_2_0_down %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> slim_results_GO_results_injury_genes_w_2_0_down_edited

# Merge GO_results with GO_slim to add GOTerms column
slim_results_GO_results_injury_genes_w_2_0_down_edited %>%
  full_join(GO_results_injury_genes_w_2_0_down, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) -> GO_results_injury_genes_w_2_0_down_all

GO_results_injury_genes_w_2_0_up_all %>% 
  mutate(Contrast = "Wounded 2h vs Wounded 0h", Direction = "Upregulated") -> GO_results_injury_genes_w_2_0_up_all

GO_results_injury_genes_w_2_0_down_all %>% 
  mutate(Contrast = "Wounded 2h vs Wounded 0h", Direction = "Downregulated") -> GO_results_injury_genes_w_2_0_down_all

# Combine up and down regulated results
GO_results_genes_by_hour_w_2_0_all <- bind_rows(GO_results_injury_genes_w_2_0_up_all, GO_results_injury_genes_w_2_0_down_all)

#write_csv(GO_results_genes_by_hour_w_2_0_all, "GO_results_genes_by_hour_w_2_0_all.csv")
```


## injury_genes_w_4_0
```{r}
injury_genes %>% 
  dplyr::rename(gene = Row.names) %>% 
  filter(contrast == "Wounded 4h vs 0h (injury-filtered)") %>% 
  filter(log2FoldChange > 0) -> injury_genes_w_4_0_up

injury_genes %>%
  dplyr::rename(gene = Row.names) %>% 
  filter(contrast == "Wounded 4h vs 0h (injury-filtered)") %>% 
  filter(log2FoldChange < 0) -> injury_genes_w_4_0_down

#Upregulated 

injury_genes_w_4_0_up_list <- unique(injury_genes_w_4_0_up$gene)

## Build topGO geneList
geneList <- as.integer(allgenes %in% injury_genes_w_4_0_up_list)
names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_injury_genes_w_4_0_up <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_injury_genes_w_4_0_up$weightFisher <- as.numeric(GO_results_injury_genes_w_4_0_up$weightFisher)
GO_results_injury_genes_w_4_0_up <- GO_results_injury_genes_w_4_0_up[GO_results_injury_genes_w_4_0_up$weightFisher < 0.05, ]

str(GO_results_injury_genes_w_4_0_up) # 18 terms

# Downregulated

injury_genes_w_4_0_down_list <- unique(injury_genes_w_4_0_down$gene)

## Build topGO geneList
geneList <- as.integer(allgenes %in% injury_genes_w_4_0_down_list)
names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_injury_genes_w_4_0_down <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_injury_genes_w_4_0_down$weightFisher <- as.numeric(GO_results_injury_genes_w_4_0_down$weightFisher)
GO_results_injury_genes_w_4_0_down <- GO_results_injury_genes_w_4_0_down[GO_results_injury_genes_w_4_0_down$weightFisher < 0.05, ]

str(GO_results_injury_genes_w_4_0_down) # 31 terms
```


### Goslim
```{r}
# Upregulated

#change Class types to ontology so it matches GOslim
GO_results_injury_genes_w_4_0_up %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_injury_genes_w_4_0_up

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_injury_genes_w_2_4_up <- GO_results_injury_genes_w_4_0_up$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_injury_genes_w_4_0_up %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_injury_genes_w_4_0_up <- bind_rows(all_slim_results)

slim_results_GO_results_injury_genes_w_4_0_up %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> slim_results_GO_results_injury_genes_w_4_0_up_edited

# Merge GO_results with GO_slim to add GOTerms column
slim_results_GO_results_injury_genes_w_4_0_up_edited %>%
  full_join(GO_results_injury_genes_w_4_0_up, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) -> GO_results_injury_genes_w_4_0_up_all

# Downregulated 

#change Class types to ontology so it matches GOslim
GO_results_injury_genes_w_4_0_down %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_injury_genes_w_4_0_down

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_injury_genes_w_4_0_down <- GO_results_injury_genes_w_4_0_down$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_injury_genes_w_4_0_down %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_injury_genes_w_4_0_down <- bind_rows(all_slim_results)

slim_results_GO_results_injury_genes_w_4_0_down %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> slim_results_GO_results_injury_genes_w_4_0_down_edited

# Merge GO_results with GO_slim to add GOTerms column
slim_results_GO_results_injury_genes_w_4_0_down_edited %>%
  full_join(GO_results_injury_genes_w_4_0_down, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) -> GO_results_injury_genes_w_4_0_down_all

GO_results_injury_genes_w_4_0_up_all %>% 
  mutate(Contrast = "Wounded 4h vs Wounded 0h", Direction = "Upregulated") -> GO_results_injury_genes_w_4_0_up_all

GO_results_injury_genes_w_4_0_down_all %>% 
  mutate(Contrast = "Wounded 4h vs Wounded 0h", Direction = "Downregulated") -> GO_results_injury_genes_w_4_0_down_all

# Combine up and down regulated results
GO_results_genes_by_hour_w_4_0_all <- bind_rows(GO_results_injury_genes_w_4_0_up_all, GO_results_injury_genes_w_4_0_down_all)

#write_csv(GO_results_genes_by_hour_w_4_0_all, "GO_results_genes_by_hour_w_4_0_all.csv")

str(GO_results_genes_by_hour_w_4_0_all)
colnames(GO_results_genes_by_hour_w_4_0_all)
```



# Bubble Plot
```{r}
GO_results_genes_by_hour_w_1_0_all$TimePoint <- "1h"
GO_results_genes_by_hour_w_2_0_all$TimePoint <- "2h"
GO_results_genes_by_hour_w_4_0_all$TimePoint <- "4h"

# Combine all
go_combined <- rbind(GO_results_genes_by_hour_w_1_0_all, GO_results_genes_by_hour_w_2_0_all, GO_results_genes_by_hour_w_4_0_all)

go_combined %>% 
  group_by(Ontology, Contrast, Direction) %>% 
  top_n(3, Significant) %>% 
  ggplot(., aes(x = reorder(Term, Significant), y = TimePoint)) +
  geom_point(aes(size = Significant, col = Direction)) +
  theme_classic() +
  labs(x = NULL, y=NULL, size = "Genes") +
  facet_wrap(~Ontology, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue")) +
  coord_flip() + 
  theme(text = element_text(size = 13))  +
  scale_size(range = c(1, 5)) 
#ggsave("GO_bubble_plot_genes_by_hour.png", width = 6, height = 8)
```




## genes_by_hour_w4_c4

```{r}
genes_by_hour %>% 
  filter(group == "Wounded_4 vs Control_4") %>% 
  filter(log2FoldChange > 0) -> genes_by_hour_w4_c4_up

genes_by_hour %>% 
  filter(group == "Wounded_4 vs Control_4") %>% 
  filter(log2FoldChange < 0) -> genes_by_hour_w4_c4_down

# Up regulated 
genes_by_hour_w4_c4_up_list <- unique(genes_by_hour_w4_c4_up$Row.names)

## Build topGO geneList
geneList <- as.integer(allgenes %in% genes_by_hour_w4_c4_up_list)
names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_genes_by_hour_w4_c4_up <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_genes_by_hour_w4_c4_up$weightFisher <- as.numeric(GO_results_genes_by_hour_w4_c4_up$weightFisher)
GO_results_genes_by_hour_w4_c4_up <- GO_results_genes_by_hour_w4_c4_up[GO_results_genes_by_hour_w4_c4_up$weightFisher < 0.05, ]

str(GO_results_genes_by_hour_w4_c4_up) # 16 terms

# down regulated 
genes_by_hour_w4_c4_down_list <- unique(genes_by_hour_w4_c4_down$Row.names)

## Build topGO geneList
geneList <- as.integer(allgenes %in% genes_by_hour_w4_c4_down_list)
names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_genes_by_hour_w4_c4_down <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_genes_by_hour_w4_c4_down$weightFisher <- as.numeric(GO_results_genes_by_hour_w4_c4_down$weightFisher)
GO_results_genes_by_hour_w4_c4_down <- GO_results_genes_by_hour_w4_c4_down[GO_results_genes_by_hour_w4_c4_down$weightFisher < 0.05, ]

str(GO_results_genes_by_hour_w4_c4_down) # 21 terms
```


Goslim
```{r}
# Upregulated

#change Class types to ontology so it matches GOslim
GO_results_genes_by_hour_w4_c4_up %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_genes_by_hour_w4_c4_up

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_genes_by_hour_w4_c4_up <- GO_results_genes_by_hour_w4_c4_up$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_genes_by_hour_w4_c4_up %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_genes_by_hour_w4_c4_up <- bind_rows(all_slim_results)

slim_results_GO_results_genes_by_hour_w4_c4_up %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> GO_slim_slim_results_GO_results_genes_by_hour_w4_c4_up_edited

# Merge GO_results with GO_slim to add GOTerms column
GO_slim_slim_results_GO_results_genes_by_hour_w4_c4_up_edited %>%
  full_join(GO_results_genes_by_hour_w4_c4_up, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) -> GO_results_genes_by_hour_w4_c4_up_all
# Downregulated 

#change Class types to ontology so it matches GOslim
GO_results_genes_by_hour_w4_c4_down %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_genes_by_hour_w4_c4_down

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_genes_by_hour_w4_c4_down <- GO_results_genes_by_hour_w4_c4_down$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_genes_by_hour_w4_c4_down %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_genes_by_hour_w4_c4_down <- bind_rows(all_slim_results)

slim_results_GO_results_genes_by_hour_w4_c4_down %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> GO_slim_results_GO_results_genes_by_hour_w4_c4_down_edited

# Merge GO_results with GO_slim to add GOTerms column
GO_slim_results_GO_results_genes_by_hour_w4_c4_down_edited %>%
  full_join(GO_results_genes_by_hour_w4_c4_down, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) -> GO_results_genes_by_hour_w4_c4_down_all

GO_results_genes_by_hour_w4_c4_up_all %>% 
  mutate(Contrast = "Wounded 4h vs Control 4h", Direction = "Upregulated") -> GO_results_genes_by_hour_w4_c4_up_all

GO_results_genes_by_hour_w4_c4_down_all %>% 
  mutate(Contrast = "Wounded 4h vs Control 4h", Direction = "Downregulated") -> GO_results_genes_by_hour_w4_c4_down_all

# Combine up and down regulated results
GO_results_genes_by_hour_w4_c4_all <- bind_rows(GO_results_genes_by_hour_w4_c4_up_all, GO_results_genes_by_hour_w4_c4_down_all)

#write_csv(GO_results_genes_by_hour_w4_c4_all, "GO_results_genes_by_hour_w4_c4_all.csv")
```







# Gene Clusters
```{r}
geneclusters <- read_csv("../3_Results/cluster_with_intersection_info.csv")

geneclusters %>% 
  filter(cluster == 15) %>% 
  mutate(Cluster = "Cluster_1") -> cluster_1_genes

geneclusters %>% 
  filter(cluster == 21) %>% 
  mutate(Cluster = "Cluster_2") -> cluster_2_genes

geneclusters %>% 
  filter(cluster == 22) %>% 
  mutate(Cluster = "Cluster_3") -> cluster_3_genes

geneclusters %>%
  filter(cluster == 24) %>% 
  mutate(Cluster = "Cluster_4") -> cluster_4_genes
```
## cluster 1
```{r}
cluster_1_genes_list <- unique(cluster_1_genes$genes)

## Build topGO geneList
geneList <- as.integer(allgenes %in% cluster_1_genes_list)

names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_cluster_1 <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_cluster_1$weightFisher <- as.numeric(GO_results_cluster_1$weightFisher)
GO_results_cluster_1 <- GO_results_cluster_1[GO_results_cluster_1$weightFisher < 0.05, ]

str(GO_results_cluster_1) # 23 terms
```

### Goslim
```{r}
#change Class types to ontology so it matches GOslim
GO_results_cluster_1 %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_cluster_1

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_cluster_1 <- GO_results_cluster_1$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_cluster_1 %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_cluster_1 <- bind_rows(all_slim_results)

slim_results_GO_results_cluster_1 %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> slim_results_GO_results_cluster_1_edited

# Merge GO_results with GO_slim to add GOTerms column
slim_results_GO_results_cluster_1_edited %>%
  full_join(GO_results_cluster_1, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) %>% 
  mutate(Cluster = "Cluster_1") -> GO_results_cluster_1_all
```


## cluster 2
```{r}
cluster_2_genes_list <- unique(cluster_2_genes$genes)

## Build topGO geneList
geneList <- as.integer(allgenes %in% cluster_2_genes_list)

names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_cluster_2 <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_cluster_2$weightFisher <- as.numeric(GO_results_cluster_2$weightFisher)
GO_results_cluster_2 <- GO_results_cluster_2[GO_results_cluster_2$weightFisher < 0.05, ]

str(GO_results_cluster_2) # 16 terms
```

### Goslim
```{r}
#change Class types to ontology so it matches GOslim
GO_results_cluster_2 %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_cluster_2

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_cluster_2 <- GO_results_cluster_2$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_cluster_2 %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_cluster_2 <- bind_rows(all_slim_results)

slim_results_GO_results_cluster_2 %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> slim_results_GO_results_cluster_2_edited

# Merge GO_results with GO_slim to add GOTerms column
slim_results_GO_results_cluster_2_edited %>%
  full_join(GO_results_cluster_2, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) %>% 
  mutate(Cluster = "Cluster_2") -> GO_results_cluster_2_all
```


## cluster 3
```{r}
cluster_3_genes_list <- unique(cluster_3_genes$genes)

## Build topGO geneList
geneList <- as.integer(allgenes %in% cluster_3_genes_list)

names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_cluster_3 <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_cluster_3$weightFisher <- as.numeric(GO_results_cluster_3$weightFisher)
GO_results_cluster_3 <- GO_results_cluster_3[GO_results_cluster_3$weightFisher < 0.05, ]

str(GO_results_cluster_3) # 17 terms
```

### Goslim
```{r}
#change Class types to ontology so it matches GOslim
GO_results_cluster_3 %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_cluster_3

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_cluster_3 <- GO_results_cluster_3$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_cluster_3 %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_cluster_3 <- bind_rows(all_slim_results)

slim_results_GO_results_cluster_3 %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> slim_results_GO_results_cluster_3_edited

# Merge GO_results with GO_slim to add GOTerms column
slim_results_GO_results_cluster_3_edited %>%
  full_join(GO_results_cluster_3, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) %>% 
  mutate(Cluster = "Cluster_3") -> GO_results_cluster_3_all
```


## cluster 4

```{r}
cluster_4_genes_list <- unique(cluster_4_genes$genes)

## Build topGO geneList
geneList <- as.integer(allgenes %in% cluster_4_genes_list)

names(geneList) <- allgenes
str(geneList)
sum(names(geneList) %in% names(geneID2GO)) #11737

## Build topGOdata objects
GOdata_BP <- new("topGOdata",
                 ontology = "BP",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_CC <- new("topGOdata",
                 ontology = "CC",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)

GOdata_MF <- new("topGOdata",
                 ontology = "MF",
                 allGenes = geneList,
                 geneSelectionFun = function(x) x == 1,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO,
                 nodeSize = 5)


## Run enrichment tests
resultTopGO.BP <- runTest(GOdata_BP,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.CC <- runTest(GOdata_CC,
                          algorithm = "weight01",
                          statistic = "fisher")

resultTopGO.MF <- runTest(GOdata_MF,
                          algorithm = "weight01",
                          statistic = "fisher")


## Extract genes per GO term
allBPGO <- genesInTerm(GOdata_BP)
allCCGO <- genesInTerm(GOdata_CC)
allMFGO <- genesInTerm(GOdata_MF)

BP_ANNOTATION <- lapply(allBPGO, function(x) x[x %in% DEG_genes])
CC_ANNOTATION <- lapply(allCCGO, function(x) x[x %in% DEG_genes])
MF_ANNOTATION <- lapply(allMFGO, function(x) x[x %in% DEG_genes])


# Create results tables (numChar parameter controls term description length)
BP_Res <- GenTable(GOdata_BP, weightFisher = resultTopGO.BP, topNodes = length(allBPGO), numChar = 1000)
BP_Res$Class <- "Biological_Process"
BP_Res$GeneList <- vapply(BP_Res$GO.ID, function(go) {
  str_c(BP_ANNOTATION[[go]], collapse = ",")
}, character(1))

CC_Res <- GenTable(GOdata_CC, weightFisher = resultTopGO.CC, topNodes = length(allCCGO), numChar = 1000)
CC_Res$Class <- "Cellular_Component"
CC_Res$GeneList <- vapply(CC_Res$GO.ID, function(go) {
  str_c(CC_ANNOTATION[[go]], collapse = ",")
}, character(1))

MF_Res <- GenTable(GOdata_MF, weightFisher = resultTopGO.MF, topNodes = length(allMFGO), numChar = 1000)
MF_Res$Class <- "Molecular_Function"
MF_Res$GeneList <- vapply(MF_Res$GO.ID, function(go) {
  str_c(MF_ANNOTATION[[go]], collapse = ",")
}, character(1))


## Combine and filter results
GO_results_cluster_4 <- bind_rows(MF_Res, CC_Res, BP_Res)
GO_results_cluster_4$weightFisher <- as.numeric(GO_results_cluster_4$weightFisher)
GO_results_cluster_4 <- GO_results_cluster_4[GO_results_cluster_4$weightFisher < 0.05, ]

str(GO_results_cluster_4) # 19 terms
```

### Goslim
```{r}
#change Class types to ontology so it matches GOslim
GO_results_cluster_4 %>% 
  mutate(Ontology = case_when(
    Class == "Biological_Process" ~ "BP",
    Class == "Molecular_Function" ~ "MF",
    Class == "Cellular_Component" ~ "CC"
  )) -> GO_results_cluster_4

# Load the GO Slim file
slim <- getOBOCollection("../../AcerCCC/gene_expression/goslim_generic.obo") #download here (https://release.geneontology.org/2024-09-08/ontology/subsets/index.html#:~:text=281%20KB-,goslim_generic.obo,-117%20KB)

slim_ids <- GSEABase::ids(slim) # gets slim IDs

# Helper function: map slim → detailed GO terms
goWide <- function(slim_id, enriched_go_terms, GO_ontology) {
  GOTERM <- getAnnMap("TERM", "GO")
  terms <- mget(slim_ids, GOTERM, ifnotfound = NA)
  terms <- terms[vapply(terms, Ontology, character(1)) == GO_ontology]
  slim_ids_o <- names(terms)
  
  OFFSPRING <- switch(GO_ontology,
    MF = getAnnMap("MFOFFSPRING", "GO"),
    BP = getAnnMap("BPOFFSPRING", "GO"),
    CC = getAnnMap("CCOFFSPRING", "GO"),
    stop("GO_ontology must be 'MF', 'BP', or 'CC'")
  )
  
  # Get all offspring (detailed terms) of this slim term
  offspring <- mget(slim_id, OFFSPRING, ifnotfound = NA)[[1]]
  
  if (is.na(offspring[1])) {
    return(character(0))
  }
  
  # Find which of the enriched GO terms are offspring of this slim term
  # Include the slim term itself if it's in the enriched list
  matching_terms <- enriched_go_terms[enriched_go_terms %in% c(slim_id, offspring)]
  
  return(matching_terms)
}

# Run GOslim for each ontology
all_slim_results <- list()
GO_IDs_GO_results_cluster_4 <- GO_results_cluster_4$GO.ID
GOTERM <- getAnnMap("TERM", "GO")

for (ontology in c("BP", "MF", "CC")) {
  message("Processing ontology: ", ontology)
  
  ontology_terms <- GO_results_cluster_4 %>% filter(Ontology == ontology)
  
  if (nrow(ontology_terms) > 0) {
    tmp <- goSlim(GOCollection(ontology_terms$GO.ID), slim, ontology) %>%
      rownames_to_column("SlimID") %>%
      mutate(Ontology = ontology) %>%
      filter(Count > 0)
    
    # Add enriched GO terms that map to each slim category
    tmp$EnrichedGOIDs <- sapply(tmp$SlimID, function(id) {
      terms <- goWide(id, ontology_terms$GO.ID, ontology)
      paste(terms, collapse = ", ")
    })
    
    # Add the descriptions of those enriched GO terms
    tmp$EnrichedGOTerms <- sapply(tmp$SlimID, function(id) {
      go_ids <- goWide(id, ontology_terms$GO.ID, ontology)
      if (length(go_ids) == 0) return("")
      
      descriptions <- sapply(go_ids, function(goid) {
        # Get the term description from your GO_results
        term_desc <- ontology_terms$Term[ontology_terms$GO.ID == goid]
        if (length(term_desc) > 0) return(term_desc[1])
        return(goid)
      })
      paste(descriptions, collapse = "; ")
    })
    
    all_slim_results[[ontology]] <- tmp
  }
}


# Combine all GOslim results
slim_results_GO_results_cluster_4 <- bind_rows(all_slim_results)

slim_results_GO_results_cluster_4 %>% 
  dplyr::rename(ParentTermID = SlimID, ParentTermDescription = Term) %>% 
  dplyr::select(!EnrichedGOTerms) %>% 
  separate_rows(EnrichedGOIDs, sep = ", ") -> slim_results_GO_results_cluster_4_edited

# Merge GO_results with GO_slim to add GOTerms column
slim_results_GO_results_cluster_4_edited %>%
  full_join(GO_results_cluster_4, by = c("EnrichedGOIDs" = "GO.ID", "Ontology")) %>% 
  mutate(Cluster = "Cluster_4") -> GO_results_cluster_4_all
```


## Combine all cluster results
```{r}
GO_results_clusters_all <- bind_rows(GO_results_cluster_1_all,
                                     GO_results_cluster_2_all,
                                     GO_results_cluster_3_all,
                                     GO_results_cluster_4_all)

#write_csv(GO_results_clusters_all, "GOenrichment_clusters_all_results.csv")

GO_results_clusters_all %>% 
group_by(ParentTermID) %>%
  mutate(
    n_clusters = n_distinct(Cluster),
    Assignment = if_else(n_clusters > 1, "Shared", "Unique")
  ) %>%
  ungroup() %>% 
  drop_na() %>% 
  dplyr::select(ParentTermID:Ontology, Cluster, Assignment) %>% 
  filter(Ontology != "CC") %>% 
  distinct() #%>% 
  #write_csv("GOenrichment_clusters_summary_shared_unique_GOslim_terms.csv")

# Plot of top GO terms
top_terms <- GO_results_clusters_all %>%
  dplyr::select(Ontology:Cluster) %>% 
  group_by(Ontology, Cluster) %>%
  arrange(weightFisher) %>%
  filter(Ontology != "CC") %>% 
  slice_head(n = 5) %>%  # Top 5 per ontology
  ungroup() %>%
  mutate(
    GeneRatio = Significant / Annotated,  # Calculate gene ratio
    NegLog10P = -log10(weightFisher),
    Term = fct_reorder(Term, NegLog10P)  # Order by significance
  )

# Create dot plot
top_terms %>% 
    filter(Ontology == "BP") %>% 
ggplot(., aes(x = GeneRatio, y = fct_reorder(Term, GeneRatio))) +
  geom_point(aes(size = Significant, color = NegLog10P)) +
  scale_color_gradient(low = "blue", high = "red", 
                       name = "-log10(P-value)") +
  scale_size_continuous(name = "Gene Count", range = c(1, 8)) +
  facet_grid(Ontology ~ ., scales = "free_y", space = "free_y") +
  labs(x = "Gene Ratio",
       y = "") +
  theme_bw() +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 14),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  ) +
  facet_wrap(~Cluster, scales = "free", ncol = 1)

```

